### 2012-11-15 doc/progress_plan/report.tex
Finished rewriting background and started rewriting similar systems, also working on 'what I've done' but

\subsection{Literature Review}

In science fiction literature and films, Artificial Intelligence (AI) and the
ability of machines to automatically process and understand human language is
almost always taken for granted. The current state of AI is a long way behind these
fantastic visions. Dale(2000) comments that ``Even the most ardent exponent of
artificial intelligence research would have to admit that the likes of HAL in
Kubrick's 2001: A Space Odyssey remain firmly in the realms of science
fiction\cite{dale2000handbook}".

@@

Turing(1950) proposed a test for determining whether a machine could be
considered intelligent or not\cite{turing1950computing}. Turing's test is based
upon whether a computer can communicate in a natural language proficiently
enough to deceive a human into thinking that the machine is also human. A
machine able to pass such a test would need to possess the ability to represent
and learn from knowledge, to be able to reason about what it knows and to be
able to process natural language\cite{russell2010artificial}. Partridge will
also need to be able to process natural language and learn from any
observations it makes. It is reasonable to consider Partridge's backend system
to be an Artificial Intelligence problem. Primarily, Patridge can be considered
a Natural Language Processing problem.

@@

Natural Language Processing (NLP) is still a relatively unexplored discipline
and as such is a very active area of research and development within the AI
community\cite{liddy2001natural}. NLP enables the automated extraction of
meaningful information from texts written in human languages such as English or
French. NLP has already been applied to many text classification and data
extraction problems. Studies have been carried out in the detection of emotions
in suicide notes\cite{citeulike:11077287}, the classification of a web page's
genre on the World Wide Web\cite{citeulike:11288938} and emotional polarity
(determining the positive or negative implications of) of a phrase or
sentence\cite{Wilson05Polarity}. Liddy(2001) defines Natural Language
Processing as:

\begin{quotation} 
A theoretically motivated range of computational techniques for analyzing and
representing naturally occurring texts at one or more levels of linguistic
analysis for the purpose of achieving human-like language processing for a
range of tasks or applications \it{(Ibid)}.  
\end{quotation}

In the case of Partridge, scientific papers, constituting the naturally occuring
texts, are processed at sentence-by-sentence and word-by-word levels of
linguistics and represented in the form of Extended Markup Language (XML)
documents, a format that is both human and machine readable. This information
is then used for the purpose of classifying and searching papers in a
human-like way.

It is therefore necessary to define what a 'human-like way' of processing
scientific papers.  Krug (2005), suggests that when browsing the internet,
humans find it much easier to locate specific information within a labelled and
logically structured document than one that is provided as a single text
entity\cite{Krug:2005:DMM:1051204}. In order to help humans to find relevant
research papers, Partridge will represent all papers in its repository in a
logical hierarchical structure. This will facilitate better information
retrieval for keyword searches and enable more advanced NLP techniques.
Partridge's document storage format should therefore be standardised to provide
a uniform way of processing each document.

@@

Soldatova and Liakata(2007) proposed a methodology for storing the Core
Information about Scientific Papers (CISP) as a way to formally represent
scientific concepts that should be present in the articles in a logical
ontology\cite{soldatova2007ontology}. They then proceed to define a schema for
their CISP ontology that defines the Core Scientific Concepts (CoreSC) as part
of the XML document itself\cite{liakata2008guidelines}.

It is proposed that Partridge uses the CoreSC schema for document storage
to provide a standard way to access and process the papers and thus
eliminate dealing with multiple documents during its learning phase. 
Using CoreSC also offers the advantage of making Partridge compatible
with the ART corpus. This is a set of physical chemistry and biochemistry
research papers that have been pre-processed and already annotated using CISP
concepts\cite{liakataART}. This would provide Partridge with a set of papers
that could be used as a training set for classification tasks and could be used
for information retrieval tests from early on in the project.

Liakata et al. (2012) describe a system for automatically processing and
categorising sentences in a research paper according to their respective CoreSC
element\cite{citeulike:10444769}. SAPIENTA
(\url{http://www.sapientaproject.com}) was trained using the ART corpus and can
achieve promising accuracies when categorising sentences (\emph{Ibid}). With
the authors' permissions, Partridge will make use of SAPIENTA when introducing
new papers into its repository. It is hoped that preclassifying papers with
SAPIENTA will make the development of Partridge more manageable within the
timeframe available and will also allow more effort to be put into other
classification problems rather than re-implementing a SAPIENTA clone. Any
modifications that have to be made to SAPIENTA to integrate it into Partridge
will be submitted back to the original authors to assist them in their research.

After pre-processing by SAPIENTA is complete, Partridge will need to do some
classification work of its own to determine the area of study (i.e. Physics,
Chemistry, Biology), type of paper (Literature review, Case Study, Experiment)
and the polarity of the result (positive, negative, inconclusive). These tasks
involve implementing existing AI techniques and applying them specifically to
the task of text classification. To avoid the time consuming task of
re-implementing these methods, it was decided that a library should be used
instead. Different NLP libraries and their properies are discussed below in
Section \ref{sec:libschoice}.

@@

There are already many existing systems for finding and filtering information
on the World Wide Web. Search engines are very useful for information retrieval
in the very large and generalised search domain of the Internet. Most people
have heard of Google (\url{http://wwww.google.com}), Yahoo
(\url{http://www.yahoo.co.uk}), Bing (\url{http://www.bing.com}) and Ask
(\url{http://www.ask.com}). There are many more similar systems available for
free general use across the internet.  They all present very similar user
interfaces (as shown in Figure \ref{fig:search_interfaces}) in which users are
asked to supply keywords that might be linked to relevant documents and the
search engine returns a list of Uniform Resource Locators (URLs) that they
consider to match the user's query.

\begin{figure}[!ht]
        \centering
        \begin{subfigure}[b]{0.50\textwidth}
                \centering
                \includegraphics[width=\textwidth]{images/ask_front.png}
                \caption{Ask.com}
                \label{fig:ask_interface}
        \end{subfigure}%
        \begin{subfigure}[b]{0.50\textwidth}
                \centering
                \includegraphics[width=\textwidth]{images/bing_front.png}
                \caption{Bing.com}
                \label{fig:bing_interface}
        \end{subfigure}\\
        \begin{subfigure}[b]{0.50\textwidth}
                \centering
                \includegraphics[width=\textwidth]{images/google_front.png}
                \caption{Google.com}
                \label{fig:google_interface}
        \end{subfigure}%
        \begin{subfigure}[b]{0.50\textwidth}
                \centering
                \includegraphics[width=\textwidth]{images/yahoo_front.png}
                \caption{Yahoo.com}
                \label{fig:yahoo_interface}
        \end{subfigure}%
        \caption{4 popular search engine interfaces}\label{fig:animals}
        \label{fig:search_interfaces}
\end{figure}

Search engines are helpful in locating pages and websites within the World Wide
Web. Unfortunately, the problem space they deal with is usually too big for
them to find scientific papers and journals given a set of keywords. Internet
search engines index a huge proportion of irrelevant information compared to
useful information\cite{Berghel1997}, and as a result, even relatively specific
queries such as ``effects of gravity on rockets" yield millions of results (as
shown in Figure \ref{fig:rocket_results}).

@@

Partridge offers an advantage over these mechanisms as it will specifically
index research papers rather than attempting to index the whole Internet.
This means that there should be a higher proportion of useful information as
output compared to the output of an Internet Search Engine.

\subsubsection{Scientific Paper Search Engines}
There are also a number of search and indexing systems that specifically look
for scientific papers as opposed to web pages. One of the most publicised and
well known paper search systems is Google Scholar
(\url{http://scholar.google.com}).  As can be seen in Figure
\ref{fig:scholar_basic}, This is an adaptation of Google's general search
engine (discussed above) to specifically index and search scientific papers.
Google also offers advanced query options specific to Scholar that allow
searching by author, year and for words that occur only in the document title
as shown in figure \ref{fig:scholar_advanced}. Whilst this does deal with the
problem of `information overload' and provides even more fine control over the
information returned from searches,  the user is still required to have a very
good idea of what they are looking for in terms of keywords and/or specific
authors. It is possible that a user would not know what they are looking for
until they've seen it. Even if the user has a set of keywords to search for,
they can only search the title of the paper or the content as a whole. This
means that users who want to find a particular phrase within a CoreSC part of
the paper (e.g. only look for this phrase in the `Result' section of the
paper) are unable to get results at their desired level of detail.

@@

Partridge will provide the option to filter papers by subject and it is hoped
that the system will also provide user-specific recommendations by profiling
them through their reading history. This will make it easier for users to find
relevant papers without knowing exactly which keywords they need to search for.
Partridge will also offer facilitate searching for keywords within a specific
CoreSC section by making use of Liakata et al's SAPIENTA project for
classifying each sentence of paper as it is added to the repository.

@@

These systems have the ability to make recommendations to the user without
requiring specific keywords or search terms. They do this by learning the
user's profile and taking into account the preferences of their 'friends' and
their browsing history. However, the above-named systems do not take into
account the content of the paper or book. They only deal with metadata as can
be seen in Figure \ref{fig:social_searches}. This means that important
discriminatory information that could be contained within the actual document
content is overlooked completely. Partridge will analyse documents on a
sentence-by-sentence and possibly word-by-word basis, thereby taking account of
any embedded information that could be missed by these social metadata systems.

@@

A prototype script was written using a Python PDF extraction library called
PDFMiner (\url{http://www.unixuser.org/~euske/python/pdfminer/index.html}).
This toolkit already contains some heuristics about how to extract text from
PDF documents in a sane way\cite{pdfminer}. The script then used the NLTK
library to split the text into sentences and save the document in a CoreSC
compatible format for processing by SAPIENTA.

The script had limited success due to the high variation of formatting within
scientific papers. It was suggested that PDFX (\url{http://pdfx.cs.man.ac.uk/})
a free service hosted by the university of manchester, could be used instead of
PDFMiner. PDFX's main advantage is that it uses a machine learning system to
guess PDF document structures more accurately than through the use of simple
Heuristics. With an adapted backend that sends PDFs for analysis using PDFX and
then splits the results using NLTK, the conversion script now has a much higher
success rate and will be used as a preprocessor for adding PDF  documents to
Partridge.

>>>>

\subsection{Background}

\subsubsection{Artificial Intelligence}
In science fiction literature and films, Artificial Intelligence (AI) and the
ability of machines to automatically process and understand human language is
almost always taken for granted. The current state of AI is a long way behind these
fantastic visions. Dale(2000) comments that ``Even the most ardent exponent of
artificial intelligence research would have to admit that the likes of HAL in
Kubrick's 2001: A Space Odyssey remain firmly in the realms of science
fiction\cite{dale2000handbook}".

@@

Turing(1950) proposed a test for determining whether a machine could be
considered intelligent or not\cite{turing1950computing}. Turing's test is based
upon whether a computer can communicate in a natural language proficiently
enough to deceive a human into thinking that the machine is also human. A
machine able to pass such a test would need to possess the ability to represent
and learn from knowledge, to be able to reason about what it knows and to be
able to process natural language\cite{russell2010artificial}.

@@

Natural Language Processing (NLP)  enables the automated extraction of
meaningful information from texts written in human languages such as French or
English. Liddy(2001) defines Natural Language Processing as:

\begin{quotation} 
A theoretically motivated range of computational techniques for analyzing and
representing naturally occurring texts at one or more levels of linguistic
analysis for the purpose of achieving human-like language processing for a
range of tasks or applications \cite{liddy2001natural}.  
\end{quotation}

Over the last 60 years, NLP has been used in a wide variety of applications
such as automated translation between languages\cite{hutchins2004first}, the
engineering of systems for querying databases using natural languages
\cite{rao2010natural} and for building `chatterbot' systems designed to
communicate with their users in a human-like way\cite{Alfonsi2006}.
More recently, NLP has been used for text classification purposes such as
classifying emotions within phrases and sentences \cite{Wilson05Polarity} and
even within suicide notes \cite{citeulike:11077287}.

@@

Liakata et al. (2012) describe a system for automatically processing and
classifying sentences in a research paper according to the scientific concept
they describe (e.g. a sentence can be categorised as a hypothesis, background
information, a method or similar)\cite{citeulike:10444769}. SAPIENTA
(\url{http://www.sapientaproject.com}) is a machine learning application,
trained using a corpus of physical chemistry and biochemistry research papers
that were pre-processed and annotated using Core Information about Scientific
Papers (CISP)\cite{LIAKATA10.644}.

CISP, Soldatova and Liakata(2007), is a way to formally represent core
scientific concepts (CoreSC), e.g. background, hypothesis, method etc., that
should be present in the articles in a logical
ontology\cite{soldatova2007ontology}. Subsequent work implements this set of
concepts as a three layered sentence based annotation scheme (CoreSC scheme)
where annotations are encoded in Extended Markup Language
(XML)\cite{LIAKATA10.644}.

@@

There are already many existing systems for finding and filtering information
on the World Wide Web. Search engines are very useful for information retrieval
in the very large and generalised search domain of the Internet. Most people
have heard of Google (\url{http://wwww.google.com}), Yahoo
(\url{http://www.yahoo.co.uk}), Bing (\url{http://www.bing.com}) and Ask
(\url{http://www.ask.com}).

@@

Despite their use as general search tools on the World Wide Web, the problem
space that search engines deal with is usually too big for them to find
scientific papers and journals given a set of keywords. Internet search engines
index a huge proportion of irrelevant information compared to useful
information\cite{Berghel1997}, and as a result, even relatively specific
queries such as ``effects of gravity on rockets" yield millions of results (as
shown in Figure \ref{fig:rocket_results}). This shortcoming has already led to
the development of search engines designed specifically to find scientific
papers.

\subsubsection{Scientific Paper Search Engines}

There are also a number of search and indexing systems that specifically look
for scientific papers as opposed to web pages. One of the most publicised and
well known paper search systems is Google Scholar
(\url{http://scholar.google.com}).  As can be seen in Figure
\ref{fig:scholar_basic}, This is an adaptation of Google's general search
engine (discussed above) to specifically index and search scientific papers.
Google also offers advanced query options specific to Scholar that allow
searching by author, year and for words that occur only in the document title
as shown in figure \ref{fig:scholar_advanced}. Whilst this does deal with the
problem of `information overload' and provides even more fine control over the
information returned from searches,  the user is still required to have a very
good idea of what they are looking for in terms of keywords and/or specific
authors. It is possible that a user would not know what they are looking for
until they've seen it. Even if the user has a set of keywords to search for,
they can only search the title of the paper or the content as a whole. This
means that users who want to find a particular phrase within a CoreSC part of
the paper (e.g. only look for this phrase in the `Result' section of the
paper) are unable to get results at their desired level of detail.

@@

@@

These systems have the ability to make recommendations to the user without
requiring specific keywords or search terms. They do this by learning the
user's profile and taking into account the preferences of their 'friends' and
their browsing history. However, the above-named systems do not take into
account the content of the paper or book. They only deal with metadata as can
be seen in Figure \ref{fig:social_searches}. This means that important
discriminatory information that could be contained within the actual document
content is overlooked completely.

@@

This was the first big challenge in the project. A prototype script was written
using a Python PDF extraction library called PDFMiner
(\url{http://www.unixuser.org/~euske/python/pdfminer/index.html}).  This
toolkit already contains some heuristics about how to extract text from PDF
documents, grouping together characters that appear very close to each other,
and separating paragraphs and headings when a larger area of whitespace is
detected\cite{pdfminer}. Despite these rules, the library still produced some
extraneous whitespace and newline characters as part of the output. A
subroutine to trim whitespace and newlines was added to the script to resolve
this problem.

The next stage was to split the text into sentences in preparation for
processing with SAPIENTA. With the assistance of the NLTK library, a sentence
splitting subroutine was implemented. This used a machine learning algorithm
that had been trained to recognise sentence boundaries to split the text. Each
sentence was then added to a CoreSC compatible XML document for processing by
SAPIENTA.

Initially, the PDF conversion subroutine had a very high error rate due to high
variation in the formatting of scientific papers. It was suggested that PDFX (\url{http://pdfx.cs.man.ac.uk/})
a free service hosted by the University of Manchester could be used instead of
PDFMiner for the initial PDF data extraction. The main advantage of PDFX over
the PDFMiner library is that is a trained machine learning system that has been
trained using a large full-text selection of articles; PDFX's classifier is
more suited to paper conversion than PDFMiner's more general heuristic
approach.

PDFX's main advantage is that it uses a machine learning system to
guess PDF document structures more accurately than through the use of simple
Heuristics. With an adapted backend that sends PDFs for analysis using PDFX and
then splits the results using NLTK, the conversion script now has a much higher
success rate and will be used as a preprocessor for adding PDF  documents to
Partridge.
